---
title: "vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(AlphaFoldMVN)
```

# $\alpha-$transformation

Suppose $\xv\in\mathbb{S}^{D-1}$. Let's start from centred log-ratio transformation:

$$
\wv_0(\xv) = \biggl(\log\biggl(\frac{x_j}{\prod_{d=1}^Dx_d^{1/D}}\biggr)\biggr)_{j=1, \dots, D}
$$

The space wher $\wv^0$ reside is

$$
\mathbb{Q}_0^{D-1} = \biggl\{\wv_0: \sum_{d=1}^D w_{0d} = 0\biggr\}
$$

$\alpha-$transformation is en generalized centred log-ratio transformation, which has

$$
\uv_\alpha(\xv) = \biggl(\frac{x_j^\alpha}{\sum_{d=1}^Dx_d^\alpha}\biggr)_{j=1, \dots, D}
$$

$$
\wv_\alpha(\xv) = \frac{D\uv_\alpha - 1 }{\alpha}
$$

As $\alpha\to 0$, the transformation above is centred log ratio. For instance

```{r}
set.seed(42)

X <- LaplacesDemon::rdirichlet(n=5000, alpha = c(1,1,1))

alphas <- c(1, 0.5, 0.25, 0, -0.25, -0.5, -1)
W_alpha <- do.call(
  rbind,
  lapply(alphas, function(a) alpha_trans(X, alpha = a) + a*2)
)

plotly::plot_ly(x = W_alpha[,1], y = W_alpha[,2], z = W_alpha[,3],
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 2, color = rep(1:7, each = nrow(X-1)),colorscale = "Viridis"))
```

The points are all on a hyperplane orthogonal to $\mathbf{1}_{D} = (1,1,\dots, 1)$. For visualization, we shift these planes by $2\alpha$ along $\mathbf{1}_{D}$. We can project them to the complementary space of $\mathbf{1}_{D}$ by the Helmart sub matrix.

$$
\yv = \zv_\alpha(\xv) = \Hv\wv_\alpha
$$

Let's further look these points:

```{r}

H <- helmert(3)

Z_alpha <- W_alpha %*% t(H)
plotly::plot_ly(x = Z_alpha[,1], y = Z_alpha[,2], z = rep(c(1,0.5, 0.25, 0, -0.25, -0.5, -1), each = nrow(X)),
                type = "scatter3d",
                mode = "markers",
                marker = list(size = 5, color = rep(1:7, each = nrow(X)),colorscale = "Viridis"))
```

Naturally , we have the inverse transformation. Now let's look at the inverse process. As we have

$$
\yv \in \mathbb{Q}_\alpha^{D-1} = \biggl\{\Hv\wv: -\frac{1}{\alpha}<w_{d}<\frac{D-1}{\alpha}, \sum_{d=1}^D w_d = 0\biggr\}
$$

we can easily see the inverse is

$$
\xv = \bigg(\frac{(1+\alpha \pi_j(\Hv'\yv))^\frac{1}{\alpha}}{\sum_{d=1}^D (1+\alpha \pi_d(\Hv'\yv))^\frac{1}{\alpha}}\biggr)_{j=1, \dots, D}
$$

Focus on the transformation between $X$ and $W$. We can see

```{r}
do.call(
  c,
  lapply(alphas, function(a) sum(which(alpha_trans_inverse(alpha_trans(X= X, alpha = a), alpha = a)$X - X > 1e-8)))
)
```

The results validate the inverse.

# $\alpha-$folding
There is one thing we need to pay attention to: the inverse function above works for only $\wv = \Hv'\yv$, where $\yv\in\mathbb{Q}_\alpha^{D-a}$. Therefore, for any points outside of the simplex in the plots, there is no interpretation for what the inverse transformation does.

However, we expect the transformation can convert the constrained data points into a regular Euclidean space, and thus we can freely model it with Gaussian distribution.

Folding is such a technique, which assumes a latent space where the observed data could be folded from the points outside of the bounds.

```{r}
p = matrix(c(0.1, 0.6, 0.3, 0.6, 0.3, 0.1, 0.3, 0.1, 0.6), nrow = 3, byrow = TRUE)
C <- c(1/3,1/3, 1/3)
v <- p - C
t <- seq(0.7, 1.42, length.out = 50)
x <- apply(v, MARGIN = 1, function(vv) t(C + vv%*%t(t)))
H <- helmert(3)
alpha = -0.5
w <- alpha_trans(x, alpha = alpha)
W <- rbind(w, alpha_trans(x, alpha = alpha, unfold = TRUE))
visualize(W = W, 
          alpha = alpha, 
          group = rep(c(1,2), each = nrow(w)), 
          shift = 0)
```

```{r}
p = matrix(c(0.1, 0.6, 0.3, 0.6, 0.3, 0.1, 0.3, 0.1, 0.6), nrow = 3, byrow = TRUE)
C <- c(1/3,1/3, 1/3)
v <- p - C
t <- seq(0.7, 1.42, length.out = 50)
x <- apply(v, MARGIN = 1, function(vv) t(C + vv%*%t(t)))
H <- helmert(3)
alpha = 0.5
w <- alpha_trans(x, alpha = alpha)
W <- rbind(w, alpha_trans(x, alpha = alpha, unfold = TRUE))
visualize(W = W, 
          alpha = alpha, 
          group = rep(c(1,2), each = nrow(w)), 
          shift = 0)
```

# $\alpha-$folding multivariate normal distribution
With the folding technique, we further assume the latent data points follow multivariate normal dsitribution.

```{r}
library(Matrix)
set.seed(1024)
true_alpha = -0.7 
mu <- runif(3-1, min = -3, max = 3)
M <- rsparsematrix(3-1, 3-1, density = 0.2, nnz = (3-1)^2)
scaling <- runif(1, min = 0, max = 1)
Sigma <- as.matrix(forceSymmetric(M %*% t(M))) * scaling
Xl <- rafmvn(n = 500, alpha = true_alpha, mean = mu, sigma = Sigma)
X <- Xl$X
folded <- Xl$folded

Y <- X
Y[folded,] <- alpha_trans(X = X[folded,], alpha = true_alpha, unfold = TRUE)
Y[!folded,] <- alpha_trans(X = X[!folded,], alpha = true_alpha, unfold = FALSE)
W <- X
W <- alpha_trans(X = X, alpha = true_alpha)
visualize(W =Y, alpha = true_alpha, group = folded+1, shift = 0)
visualize(W =W, alpha = true_alpha, group = folded+1, shift = 0)
```


# $\alpha-$folding multivariate normal density

```{r}
library(Matrix)
library(plotly)
set.seed(1024)
true_alpha = 0.7 
mu <- runif(3-1, min = -3, max = 3)
M <- rsparsematrix(3-1, 3-1, density = 0.2, nnz = (3-1)^2)
scaling <- runif(1, min = 0, max = 1)
Sigma <- as.matrix(forceSymmetric(M %*% t(M))) * scaling
gr_list <- vector("list", nb * (nb + 1) / 2)
k <- 0
for (i in 1:nb) {
  for (j in 1:(nb - i + 1)) {
    k <- k + 1
    gr_list[[k]] <- c(by * i, by * j, 1 - by * i - by * j)
  }
}
grids <- do.call(rbind, gr_list)  

comp_to_cartesian <- function(x) {
  x <- x / rowSums(x)
  cbind(x[,2] + 0.5 * x[,3], sqrt(3)/2 * x[,3])
}
cart_grids <- comp_to_cartesian(grids) 

z <- dafmvn(X = grids, alpha = true_alpha, mean = mu, sigma = Sigma, log = TRUE)

gr <- with(data.frame(x = cart_grids[,1], y = cart_grids[,2], z = z),
           akima::interp(x, y, z, duplicate = "mean", nx = 150, ny = 150))
plot_ly(x = gr$x, y = gr$y, z = gr$z) |>
  add_markers() |>
  layout(scene = list(
    xaxis = list(title = "X"),
    yaxis = list(title = "Y"),
    zaxis = list(title = "z")
  ))

df_grid <- expand.grid(x = gr$x, y = gr$y)
df_grid$z <- as.vector(gr$z)
plot_ly(x = df_grid$x, 
        y = df_grid$y, 
        z = df_grid$z,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 2,
                      color = df_grid$z,
                      colorscale = "Viridis",
                      showscale = TRUE)) |>
  layout(scene = list(
    xaxis = list(title = "X"),
    yaxis = list(title = "Y"),
    zaxis = list(title = "Z")
  ))
```


# Maximum Likelihood Estimation
Given $\alpha$, the density parameters $\mu$ and $\Sigma$ are estimated by maximum likelihood estimation.

The data points following $\alpha-$folding multivariate normal can be generated from `rafmvn()`.

```{r}
library(Matrix)
library(mvtnorm)
set.seed(42)

true_alpha <- 0.5
D = 3
N = 500
mu <- runif(D-1, min = -3, max = 3)
M <- rsparsematrix(D-1, D-1, density = 0.2, nnz = (D-1)^2)
scaling <- runif(1, min = 0, max = 1)
Sigma <- as.matrix(forceSymmetric(M %*% t(M))) * scaling
Xl <- rafmvn(n = N, alpha = true_alpha, mean = mu, sigma = Sigma)
X <- Xl$X
folded <- Xl$folded
W <- alpha_trans(X = X, alpha = true_alpha, unfold = folded)
visualize(W = W, alpha = true_alpha, group = folded + 1, shift = 0)
```

The MLE given the correct $\alpha$ is given in the following way.

```{r}
mle <- afmvn_mle(X = X, alpha = true_alpha)
MSE <- list(mse_mu = (mle$mu_hat - mu)^2, mse_sigma = (mle$Sigma_hat - Sigma)^2)
MSE
```

The esimation of $alpha$ is processed by grid search.
```{r}
alpha_r <- alpha_est(X = X, by = 0.01)
alpha_r$alpha_hat
plot(alpha_r$loglik)
```

# Simulation 1

```{r}
set.seed(42)
seeds <- sample(1:10000, size = 100)
true_alpha <- 0.7
D = 3
N = 500

mse_mu <- list()
mse_sigma <- list()

for(i in 1:100){
  set.seed(seeds[i])
  mu <- runif(D-1, min = -3, max = 3)
  M <- rsparsematrix(D-1, D-1, density = 0.2, nnz = (D-1)^2)
  scaling <- runif(1, min = 0, max = 1)
  Sigma <- as.matrix(forceSymmetric(M %*% t(M))) * scaling
  Xl <- rafmvn(n = N, alpha = true_alpha, mean = mu, sigma = Sigma)
  X <- Xl$X
  folded <- Xl$folded
  mle <- afmvn_mle(X = X, alpha = true_alpha,max.iter = 100)
  mse_mu[[i]] <- (mle$mu_hat - mu)^2
  mse_sigma[[i]] <- (mle$Sigma_hat - Sigma)^2
}

problematic <- which(sapply(mse_mu, function(mm) sum(mm)) > 0.1)
i = problematic[3]
set.seed(seeds[i])
mu <- runif(D-1, min = -3, max = 3)
M <- rsparsematrix(D-1, D-1, density = 0.2, nnz = (D-1)^2)
scaling <- runif(1, min = 0, max = 1)
Sigma <- as.matrix(forceSymmetric(M %*% t(M))) * scaling
Xl <- rafmvn(n = N, alpha = true_alpha, mean = mu, sigma = Sigma)
X <- Xl$X
folded <- Xl$folded
mle <- afmvn_mle(X = X, alpha = true_alpha)
mse_1 <- (mle$mu_hat - mu)^2
mse_2 <- (mle$Sigma_hat - Sigma)^2
W <- alpha_trans(X, alpha = true_alpha, unfold = FALSE)
W_alpha <- alpha_trans(X, alpha = true_alpha, unfold = folded)
visualize(W = rbind(W, W_alpha), alpha = true_alpha, group = folded+1, shift = rep(c(0,0.5), each = nrow(W)))
Xl2 <- rafmvn(n = N, alpha = true_alpha, mean = mle$mu_hat, sigma = mle$Sigma_hat)
X2 <- Xl2$X
folded2 <- Xl2$folded
W2 <- alpha_trans(X2, alpha = true_alpha, unfold = FALSE)
W_alpha2 <- alpha_trans(X2, alpha = true_alpha, unfold = folded)
visualize(W = rbind(W2, W_alpha2), alpha = true_alpha, group = folded+1, shift = rep(c(0,-0.5), each = nrow(W)))
```